Manuscript Code

### 2022-09-14 Manuscript Code.Rmd
Edited for EHP peer review

## Convert native id column to block column
block_preds$block_key <- substr(block_preds$native_id, 1, 15)

@@

## Merge census data with predictions
census_block$block_key <- as.character(census_block$block_key)
census_preds <- block_preds %>%
  left_join(census_block, by="block_key") %>%
  left_join(census_blockgroup, by="blockgroup_key") %>%
  ## Exclude blocks with 0 population, 0 households, or with 0 income data
  filter(bk_tot_pop_2010 != 0) %>%
  filter(households != 0) %>%
  filter(!is.na(bg_med_inc_hshld))

@@

## Exposure Disparities by Income

@@

f <- function(x) {
  r <- quantile(x, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  r
}

ggplot(census_preds_wide, aes(x=income_group, y=pnc_noscreen)) + stat_summary(fun.data = f, geom="boxplot", fill = "#F8766D") + labs(x= "Income Level", y = "UFP Concentration (pt/cm3)") + theme(legend.position='none', text = element_text(size = 20), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + expand_limits(y=0)

ggplot(census_preds_wide, aes(x=income_group, y=ma200_ir_bc1)) + stat_summary(fun.data = f, geom="boxplot", fill = "#7CAE00") + labs(x= "Income Level", y = "BC Concentration (µg/m3)") + theme(legend.position='none', text = element_text(size = 20), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + expand_limits(y=0)

ggplot(census_preds_wide, aes(x=income_group, y=no2)) + stat_summary(fun.data = f, geom="boxplot", fill = "#00BFC4") + labs(x= "Income Level", y = "NO2 Concentration (ppb)") + theme(legend.position='none', text = element_text(size = 20), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + expand_limits(y=0)

ggplot(census_preds_wide, aes(x=income_group, y=pm2.5_ug_m3)) + stat_summary(fun.data = f, geom="boxplot", fill = "#C77CFF") + labs(x= "Income Level", y = "PM2.5 Concentration (µg/m3)") + theme(legend.position='none', text = element_text(size = 20), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + expand_limits(y=0)

@@

## Exposure Disparities by Race-Ethnicity

@@

## Table 2

@@

census_preds_long %>%
  group_by(race) %>%
  summarise(pop = sum(percent * bk_tot_pop_2010 / 100),
            ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010 * percent),
            ufp_pct_diff = per_diff(ufp, 6719),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010 * percent),
            bc_pct_diff = per_diff(bc, 522),
            no2 = weighted.mean(no2, bk_tot_pop_2010 * percent),
            no2_pct_diff = per_diff(no2, 8.77),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010 * percent),
            pm2.5_pct_diff = per_diff(pm2.5, 4.56))

@@

## Exposure Disparities by Redlining Grade

@@

## Table 3

## Pollution means weighted by block population divided by number of grades block is assigned to
redlining_blocks %>%
  group_by(holc_grade) %>%
  summarise(ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010/count),
            ufp_pct_diff = per_diff(ufp, 7102),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010/count),
            bc_pct_diff = per_diff(bc, 539),
            no2 = weighted.mean(no2, bk_tot_pop_2010/count),
            no2_pct_diff = per_diff(no2, 9.3),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010/count),
            pm2.5_pct_diff = per_diff(pm2.5, 4.7))

@@

coordinates(df_resids_locs) = ~longitude+latitude
vario = variogram(resids_pnc~longitude+latitude, df_resids_locs)
plot(vario)

@@

# Centering by holc grade A gives us intercepts close to those from single predictor model for holc grade
mlr_df <- redlining_blocks %>% 
  mutate(adjusted_inc = bg_med_inc_hshld-117134,
        adjusted_race = bk_p_ethn_non_hisp_white -85.82) %>%
  pivot_longer(cols=c("ma200_ir_bc1", "no2", "pm2.5_ug_m3", "pnc_noscreen"),
               names_to="variable",
               values_to="prediction") %>% 
  group_by(variable) %>% 
  nest()

@@

mlr_output %>% 
  mutate(tidy = map(model, broom::tidy)) %>%
  unnest(tidy) %>% 
  dplyr::select(term, estimate, std.error) %>% 
  mutate(conf.lo = estimate - 1.96*std.error, 
         conf.hi = estimate + 1.96*std.error) %>% 
  mutate(estimate = ifelse(term == "adjusted_inc", estimate*10000, estimate),
          std.error = ifelse(term == "adjusted_inc", std.error*10000, std.error),
          conf.lo = ifelse(term == "adjusted_inc", conf.lo*10000, conf.lo),
          conf.hi = ifelse(term == "adjusted_inc", conf.hi*10000, conf.hi)) %>%
  mutate(estimate = ifelse(term == "adjusted_race", estimate*10, estimate),
          std.error = ifelse(term == "adjusted_race", std.error*10, std.error),
          conf.lo = ifelse(term == "adjusted_race", conf.lo*10, conf.lo),
          conf.hi = ifelse(term == "adjusted_race", conf.hi*10, conf.hi)) %>%
  kable() %>% 
  kable_styling()

fig_4$race_eth_grp <- factor(fig_4$race_eth_grp, levels = c("Non-Hispanic White", "Asian", "Hispanic", "Black"))

ggplot(data=fig_4, aes(x=race_eth_grp, y=estimate *5, ymin=conf.lo *5, ymax=conf.hi *5)) +
  geom_pointrange(aes(col=race_eth_grp)) + 
  geom_hline(aes(col=race_eth_grp), yintercept=0, linetype=2) +
  xlab("Pollutant") + ylab("Mean Difference (95% CI) per 5% Increase in Population") + 
  geom_errorbar(aes(ymin=conf.lo*5, ymax=conf.hi*5, col=race_eth_grp), width=0.5 ,cex=1) + 
  facet_wrap(~factor(variable, levels=c("UFP (pt/cm3)","BC (µg/m3)","NO2 (ppb)","PM2.5 (µg/m3)")), strip.position="left", nrow=4, scales = "free_x") +
  theme(plot.title=element_text(size=16,face="bold"),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
      axis.text.x=element_text(face="bold"),
      axis.title=element_text(size=12,face="bold"),
      strip.text.y = element_text(hjust=0,vjust = 1,angle=180,face="bold")) +
  labs(color="Race-Ethnicity") +
  coord_flip()

@@

## Income Table

@@

## Table S1

census_preds_wide %>%
  group_by(income_group) %>%
  summarise(pop = sum(bk_tot_pop_2010),
            ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010),
            ufp_pct_diff = per_diff(ufp, 5644),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010),
            bc_pct_diff = per_diff(bc, 448),
            no2 = weighted.mean(no2, bk_tot_pop_2010),
            no2_pct_diff = per_diff(no2, 7.6),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010),
            pm2.5_pct_diff = per_diff(pm2.5, 4.3))

@@

## Sensitivity Analysis

ggplot(redlining_blocks, aes(x=income_group, y=pnc_noscreen)) + stat_summary(fun.data = f, geom="boxplot", fill = "#F8766D") + labs(x= "Income Level", y = "UFP Concentration (pt/cm3)") + theme(legend.position='none', text = element_text(size = 20), axis.text.x = element_text(angle = 45, hjust=1, vjust=1)) + expand_limits(y=0)

ggplot(redlining_blocks, aes(x=income_group, y=ma200_ir_bc1)) + stat_summary(fun.data = f, geom="boxplot", fill = "#7CAE00") + labs(x= "Income Level", y = "BC Concentration (µg/m3)") + theme(legend.position='none', text = element_text(size = 20), axis.text.x = element_text(angle = 45, hjust=1, vjust=1)) + expand_limits(y=0)

ggplot(redlining_blocks, aes(x=income_group, y=no2)) + stat_summary(fun.data = f, geom="boxplot", fill = "#00BFC4") + labs(x= "Income Level", y = "NO2 Concentration (ppb)") + theme(legend.position='none', text = element_text(size = 20), axis.text.x = element_text(angle = 45, hjust=1, vjust=1)) + expand_limits(y=0)

ggplot(redlining_blocks, aes(x=income_group, y=pm2.5_ug_m3)) + stat_summary(fun.data = f, geom="boxplot", fill = "#C77CFF") + labs(x= "Income Level", y = "PM2.5 Concentration (µg/m3)") + theme(legend.position='none', text = element_text(size = 20), axis.text.x = element_text(angle = 45, hjust=1, vjust=1)) + expand_limits(y=0)

@@

redlining_blocks_long <- redlining_blocks %>%
  pivot_longer(cols = c(bk_p_race_asian, bk_p_race_black, bk_p_ethn_hisp, bk_p_ethn_non_hisp_white, bk_p_race_native, bk_p_race_pacific), names_to="race", values_to="percent")

redlining_blocks_unique %>%
  group_by(income_group) %>%
  summarise(pop = sum(bk_tot_pop_2010),
            ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010),
            ufp_pct_diff = per_diff(ufp, 5644),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010),
            bc_pct_diff = per_diff(bc, 448),
            no2 = weighted.mean(no2, bk_tot_pop_2010),
            no2_pct_diff = per_diff(no2, 7.6),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010),
            pm2.5_pct_diff = per_diff(pm2.5, 4.3))

@@

redlining_blocks_long %>%
  group_by(race) %>%
  summarise(pop = sum(percent * bk_tot_pop_2010 / 100),
            ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010 * percent),
            ufp_pct_diff = per_diff(ufp, 8309),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010 * percent),
            bc_pct_diff = per_diff(bc, 636.9),
            no2 = weighted.mean(no2, bk_tot_pop_2010 * percent),
            no2_pct_diff = per_diff(no2, 11.64),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010 * percent),
            pm2.5_pct_diff = per_diff(pm2.5, 5.038))

>>>>

## Convert native id column to block column
block_preds$block_key <- substr(block_preds$native_id, 1, 15)
block_preds$blockgroup_key <- substr(block_preds$native_id, 1, 12)

@@

## Merge census data with predictions
census_block$block_key <- as.character(census_block$block_key)
census_blockgroup$blockgroup_key <- as.character(census_blockgroup$blockgroup_key)
census_preds <- block_preds %>%
  left_join(census_block, by="block_key") %>%
  left_join(census_blockgroup, by="blockgroup_key")  %>%
  filter(households != 0) %>%
  filter(bk_tot_pop_2010 != 0) %>%
  filter(!is.na(bg_med_inc_hshld))

@@

#### Number of blocks split between redlining grades

## Table of blocks and how many times they are graded
n_occur <- data.frame(table(redlining_blocks$block_key))

## Total number of blocks in HOLC-graded area
nrow(n_occur)

## Total number of blocks split 
nrow(n_occur[n_occur$Freq > 1,])

## Number of blocks split between 3 grades
nrow(n_occur[n_occur$Freq > 2,])

## Table 2. Exposure Disparities by Income

@@

census_preds_wide %>%
  group_by(income_group) %>%
  summarise(pop = sum(bk_tot_pop_2010),
            ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010),
            ufp_pct_diff = per_diff(ufp, 5712),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010),
            bc_pct_diff = per_diff(bc, 453),
            no2 = weighted.mean(no2, bk_tot_pop_2010),
            no2_pct_diff = per_diff(no2, 7.63),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010),
            pm2.5_pct_diff = per_diff(pm2.5, 4.27))

@@

## Table 3. Exposure Disparities by Race-Ethnicity

@@

@@

census_preds_long %>%
  group_by(race) %>%
  summarise(pop = sum(percent * bk_tot_pop_2010 / 100),
            ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010 * percent),
            ufp_pct_diff = per_diff(ufp, 6678),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010 * percent),
            bc_pct_diff = per_diff(bc, 520),
            no2 = weighted.mean(no2, bk_tot_pop_2010 * percent),
            no2_pct_diff = per_diff(no2, 8.74),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010 * percent),
            pm2.5_pct_diff = per_diff(pm2.5, 4.56))

@@

## Table 4. Exposure Disparities by Redlining Grade

@@

## Pollution means weighted by block population divided by number of grades block is assigned to
redlining_blocks %>%
  group_by(holc_grade) %>%
  summarise(ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010/count),
            ufp_pct_diff = per_diff(ufp, 7150),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010/count),
            bc_pct_diff = per_diff(bc, 547),
            no2 = weighted.mean(no2, bk_tot_pop_2010/count),
            no2_pct_diff = per_diff(no2, 9.29),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010/count),
            pm2.5_pct_diff = per_diff(pm2.5, 4.75))

@@

coordinates(df_resids_locs) = ~longitude+latitude
# vario = variogram(resids_pnc~longitude+latitude, df_resids_locs)
# plot(vario)

@@

# Centering by holc grade A gives us intercepts close to those from single predictor model for holc grade
mlr_df <- redlining_blocks %>% 
  mutate(adjusted_inc = log10(bg_med_inc_hshld) - log10(117134),
        adjusted_race = bk_p_ethn_non_hisp_white -85.82) %>%
  pivot_longer(cols=c("ma200_ir_bc1", "no2", "pm2.5_ug_m3", "pnc_noscreen"),
               names_to="variable",
               values_to="prediction") %>% 
  group_by(variable) %>% 
  nest()

@@

mlr_output %>% 
  mutate(tidy = map(model, broom::tidy)) %>%
  unnest(tidy) %>% 
  dplyr::select(term, estimate, std.error) %>% 
  mutate(conf.lo = estimate - 1.96*std.error, 
         conf.hi = estimate + 1.96*std.error) %>% 
  mutate(estimate = ifelse(term == "adjusted_inc", estimate*log10(1.1), estimate),
          std.error = ifelse(term == "adjusted_inc", std.error*log10(1.1), std.error),
          conf.lo = ifelse(term == "adjusted_inc", conf.lo*log10(1.1), conf.lo),
          conf.hi = ifelse(term == "adjusted_inc", conf.hi*log10(1.1), conf.hi)) %>%
  mutate(estimate = ifelse(term == "adjusted_race", estimate*10, estimate),
          std.error = ifelse(term == "adjusted_race", std.error*10, std.error),
          conf.lo = ifelse(term == "adjusted_race", conf.lo*10, conf.lo),
          conf.hi = ifelse(term == "adjusted_race", conf.hi*10, conf.hi)) %>%
  kable() %>% 
  kable_styling()

fig_4$race_eth_grp <- factor(fig_4$race_eth_grp, levels = c("Non-Hispanic White", "Asian", "Hispanic", "Black"))
fig_4 <- fig_4 %>%
  group_by(variable) %>%
  mutate(low = -(max(abs(conf.hi*5))), high = max(conf.hi*5))

ggplot(data=fig_4, aes(x=race_eth_grp, y=estimate*5, ymin=conf.lo*5, ymax=conf.hi*5)) +
  geom_pointrange(aes(col=race_eth_grp)) + 
  geom_hline(aes(col=race_eth_grp), yintercept=0, linetype=2) +
  xlab("Pollutant") + ylab("Mean Difference (95% CI) per 5% Increase in Population") + 
  geom_errorbar(aes(ymin=conf.lo*5, ymax=conf.hi*5, col=race_eth_grp), width=0.5 ,cex=1) + 
  facet_wrap(~factor(variable, levels=c("UFP (pt/cm3)","BC (ng/m3)","NO2 (ppb)","PM2.5 (µg/m3)")), strip.position="left", nrow=4, scales = "free_x") +
  geom_blank(aes(y = low)) +
  geom_blank(aes(y = high)) +
  theme(plot.title=element_text(size=16,face="bold"),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
      axis.text.x=element_text(face="bold"),
      axis.title=element_text(size=12,face="bold"),
      strip.text.y = element_text(hjust=0,vjust = 1,angle=180,face="bold")) +
  labs(color="Race-Ethnicity") +
  coord_flip()

@@

### Sensitivity Analysis: HOLC-Graded Area Definitions

@@

## Assign A-D HOLC grades - Analysis 1
holc_grades$duplicated <- duplicated(holc_grades$BLOCKID10)
between_grades <- holc_grades[duplicated(holc_grades$BLOCKID10), ]$BLOCKID10
holc_no_dup <- holc_grades[!(holc_grades$BLOCKID10 %in% between_grades), ]

holc_no_dup$BLOCKID10 <- as.character(holc_no_dup$BLOCKID10)
redlining_blocks <- census_preds_wide %>%
  left_join(holc_no_dup, by=c("block_key"="BLOCKID10"))
redlining_blocks$count <- as.numeric(ave(redlining_blocks$block_key, redlining_blocks$block_key, FUN = length))

## Assign X HOLC grade - Analysis 1
redlining_blocks$holc_grade[is.na(redlining_blocks$holc_grade) & redlining_blocks$block_key %in% redlining_area$BLOCKID10] <- "X"
redlining_blocks <- na.omit(redlining_blocks)

## Assign A-D HOLC grades - Analysis 2
holc_areas <- read.csv("/projects/trap/transfer/kaya/area definition.csv")
holc_areas <- holc_areas[c("BLOCKID10", "holc_grade", "area")]
holc_area_per <- holc_areas %>%
    group_by(BLOCKID10) %>%
    mutate(percent = area/sum(area))

holc_area_per$BLOCKID10 <- as.character(holc_area_per$BLOCKID10)
redlining_blocks_area <- census_preds_wide %>%
  left_join(holc_area_per, by=c("block_key"="BLOCKID10"))

## Assign X HOLC grade - Analysis 2
redlining_blocks_area$holc_grade[is.na(redlining_blocks_area$holc_grade) & redlining_blocks_area$block_key %in% redlining_area$BLOCKID10] <- "X"
redlining_blocks_area <- na.omit(redlining_blocks_area)

## Analysis 1: Pollution means weighted by block population divided by number of grades block is assigned to
redlining_blocks %>%
  group_by(holc_grade) %>%
  summarise(ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010/count),
            ufp_pct_diff = per_diff(ufp, 7102),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010/count),
            bc_pct_diff = per_diff(bc, 539),
            no2 = weighted.mean(no2, bk_tot_pop_2010/count),
            no2_pct_diff = per_diff(no2, 9.3),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010/count),
            pm2.5_pct_diff = per_diff(pm2.5, 4.7))

## Analysis 2: Pollution means weighted by block population divided by number of grades block is assigned to
redlining_blocks_area %>%
  group_by(holc_grade) %>%
  summarise(ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010*percent),
            ufp_pct_diff = per_diff(ufp, 7102),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010*percent),
            bc_pct_diff = per_diff(bc, 539),
            no2 = weighted.mean(no2, bk_tot_pop_2010*percent),
            no2_pct_diff = per_diff(no2, 9.3),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010*percent),
            pm2.5_pct_diff = per_diff(pm2.5, 4.7))

@@

### Sensitivity Analysis: Results Restricted to the HOLC-Graded Area

@@

redlining_blocks_long <- redlining_blocks_unique %>%
  pivot_longer(cols = c(bk_p_race_asian, bk_p_race_black, bk_p_ethn_hisp, bk_p_ethn_non_hisp_white, bk_p_race_native, bk_p_race_pacific), names_to="race", values_to="percent")

redlining_blocks_unique %>%
  group_by(income_group) %>%
  summarise(pop = sum(bk_tot_pop_2010),
            ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010),
            ufp_pct_diff = per_diff(ufp, 7192),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010),
            bc_pct_diff = per_diff(bc, 557),
            no2 = weighted.mean(no2, bk_tot_pop_2010),
            no2_pct_diff = per_diff(no2, 9.76),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010),
            pm2.5_pct_diff = per_diff(pm2.5, 4.92))

@@

redlining_blocks_long %>%
  group_by(race) %>%
  summarise(pop = sum(percent * bk_tot_pop_2010 / 100),
            ufp = weighted.mean(pnc_noscreen, bk_tot_pop_2010 * percent),
            ufp_pct_diff = per_diff(ufp, 8232),
            bc = weighted.mean(ma200_ir_bc1, bk_tot_pop_2010 * percent),
            bc_pct_diff = per_diff(bc, 625),
            no2 = weighted.mean(no2, bk_tot_pop_2010 * percent),
            no2_pct_diff = per_diff(no2, 11.4),
            pm2.5 = weighted.mean(pm2.5_ug_m3, bk_tot_pop_2010 * percent),
            pm2.5_pct_diff = per_diff(pm2.5, 4.99))
